Mining Interaction Log Data in a Creative Arts Practice using Transition Matrices

### 2015-09-23 flux-entropy-paper/flux-entropy-paper.tex
changed ACM category
not sure how to deal with the sub sub category descriptor…
revised classification keywords again
mystery error is caused by references: hyper refs going over page
boundaries.
major edit from Ben's version
note terminology for context: rehearsal, recording, concert; many other
changes; hopefully getting closer

\pagenumbering{arabic}

@@

\title{Mining Interaction Log Data in a Creative Arts Practice using Macro Gestures and Transition Matrices}

@@

\numberofauthors{3}
\author{%
  \alignauthor{1st Author Name\\
    \affaddr{Affiliation}\\
    \affaddr{City, Country}\\
    \email{e-mail address}}\\
  \alignauthor{2nd Author Name\\
    \affaddr{Affiliation}\\
    \affaddr{City, Country}\\
    \email{e-mail address}}\\
  \alignauthor{3rd Author Name\\
    \affaddr{Affiliation}\\
    \affaddr{City, Country}\\
    \email{e-mail address}}\\
}

@@

  In this note we describe the analysis of an interaction data archive
  associated with a long-term practice-led research program
  investigating ensemble musical performance practice on touch
  screens. We use transition matrix measures to distinguish between
  sessions of different contexts, and the internal structure of these
  sessions. The data analysis proceeds by first classifying these data
  into higher level gesture types and summarising the performers'
  transitions between these gestures. We find that two
  transition-matrix measure, flux, and entropy, can be used to
  distinguish different performance types and contexts, as well as the
  internal structure of the performances in our corpus.
\end{abstract}

@@

\category{H.5.5.}{Information Interfaces and Presentation
  (e.g. HCI)}{Sound and Music Computing}{Systems}

% \category{See
%   \url{http://acm.org/about/class/1998/} for the full list of ACM
%   classifiers. This section is required.}{}{}

%\section{General comments}
% \begin{itemize}
% \item we should be clear (and careful) where we're talking about
%   groups TMs, and where we're talking about individual performers
% \item what's the terminology we're using for a ``performance''? Is it
%   ``performance'', or is that too music-specific? ``Session''?
%   Alternatives are ``activity'', or ``interaction'', but they're so
%   general\ldots This decision has implications for our other
%   terminology: participant vs performer vs musician, etc.
% \item Title: maybe ``Traces through time: making sense of longitudinal
%   open-ended interaction data'' or some such? I think that's our big
%   contribution: it takes the in-the-wild, practitioner-driven research
%   program stuff and tries to make some quantitative sense out of it
%   post-hoc.
% \end{itemize}

@@

\begin{figure*}
  \centering
  \includegraphics[width=\linewidth]{figures/three-performance-contexts}
  \caption{Collaborative musical interaction with our touch-screen
    interfaces in three different contexts: (left to right) rehearsal
    in a studio,
    performance with percussion instruments, and performance with just
    iPads.\label{fig:three-performance-contexts}}
\end{figure*}

Modern computing devices and interfaces continue to present
opportunities to support new kinds of creative expression and
collaborative interaction~\cite{Resnick:2005yu}. However, it has been
acknowledged that understanding and evaluating interfaces in creative
contexts can be challenging~\cite{Shneiderman:2007qv}. This problem is
even more pronounced in artistic \emph{performance}, where subjective
interface evaluation through questionnaires or interviews may be both
practically difficult and culturally inappropriate. When the interface
is digital, however, it is still usually trivial to log interaction in
these situations, which is useful for curation and
preservation~\cite{England:2014ys}. The question is: can this log data
be used not just as a record for posterity, but in support of a
broader creative-practice led research program?

In this note we will show how an archive of collaborative performance
data logs can be subjected to a quantitative analysis based on
transition probabilities between gestural states. We will demonstrate
that two measures on transition matrices, flux and entropy, show
significant differences between different kinds of performances, and
give insight into the internal structure of these performances.

Our archive of interaction data logs has been accumulated to support a
long-term practice-led research program into simple iPad interfaces
for group music-making similar to those described in Martin et
al~\cite{Martin:2014cr}. Over a period of more than two years (April
2013--June 2015), we collected data from group music-making sessions
where the participants used our custom-built, touch-screen apps on
Apple iPad devices. Although several different apps have been included
in our corpus, they shared a common mode of free-form touch
interaction where tapping produced short sounds and swirling or
swiping produced long sounds. Apart from these similarities, the apps
featured a variety of sound palettes, different kinds of visual
feedback, and different kinds of networked interactions (between the
iPads in an ensemble and with a central server).

In all of our 95 interaction sessions, every touch interaction---every
touch down, dragging movement and release---was time-stamped and
logged to disk for later analysis. These sessions included
formally-structured HCI experiments, where participants were
interviewed or surveyed. They also included rehearsals without an
audience and performances at arts festivals where creative goals were
paramount; in these situations surveys or interviews would have been
inappropriate and possibly disruptive. Some of the performances were
iPad-only and others used iPads together with other acoustic
percussion instruments. Although the majority of the sessions were
free-form improvisations, a number of rehearsals and performances
involved composed pieces where the musicians read and performed
percussion notation using the apps.

Some of these session contexts are are shown
Figure~\ref{fig:three-performance-contexts}) and other descriptive
statistics about the sessions are shown in Table \ref{corpus-table}.
In our corpus, the median length of sessions is 7m26s and the median
number of participants per ensemble is four. Each session record is
labelled with metadata about the session \emph{context} (performance,
rehearsal, study), session \emph{type} (improvisation or composition),
and \emph{instrumentation} (iPads-only, or iPads with acoustic
percussion). The distribution of recorded sessions by context and type
is shown in Figure \ref{fig:count-data}.

@@

\begin{table}
\centering
\begin{tabular}{l|llll}
\hline
            & Total & Min  & Median   & Max     \\ 
\hline
N           & 95    &      &          &         \\
%Record size & 306MB     &         &        &   \\
Length & 13H9M19S & 1M19S & 7M26S & 22M20S \\
Participants& 353   & 2    & 4        & 9        \\
Flux &   & 0.02239 & 0.1445 & 0.3246\\
Entropy &   & 0.7683 & 3.342 & 4.442\\          
\hline
\end{tabular}
\caption{
  Representative data about the corpus of musical touch-screen
  interaction sessions used in this study. These sessions were
  rehearsals, performances, or studies and
  participants improvised or performed compositions. Each record
  consisted of touch-data recorded in CSV format and was later analysed
  with a gestural classification system. Some participants were
  present in multiple sessions, however this data was not tracked in
  the corpus.\label{corpus-table}}
\end{table}

@@

A useful way of modelling user interfaces is to consider their
theoretical configurations as state vectors and to deploy the methods
of matrix algebra to describe transitions between these states during
an interaction sequence~\cite{Thimbleby:2001kq, Thimbleby:2004fj}. An
inverted approach is to record users' interactions and manually code
interaction states to produce a transition model. This method has been
applied to usability analysis of resource planning software for
security applications~\cite{Kannampallil:2007fp} where the transition
probability matrix of these state sequences was used to draw
conclusions about the nature of this interactions. In the creative
arts, editing commands at a computer interface to ``live code''
computer music has also been coded as states and analysed using
transition matrices~\cite{Swift:2014tya}, in both the textual and
musical dimensions, in order to arrive at conclusions regarding
artistic style.

In our corpus of iPad musical performances, logs of low-level touch
data record each performer's micro-gestures. However, it has been
established that musical performances on touch-screens can be seen in
terms of vocabularies of continuous macro gestures that unfold over
several seconds of touch interaction~\cite{Martin:2014cr}. Improvised
performances can be then thought of as a sequence of such
states~\cite{Pressing:1988uo} for each performer.

If the performers' touch interactions can be classified as belonging
to one of $m$ distinct states, then over the course of a session each
participant $X$ generates a length $N$ sequence of these interaction
states $X_n$ ($n = 1, \ldots, N$) which can be thought of as Markov
process.

If there is an (automated) classification process for identifying each
participant's state sequence from their interaction data, then we can
estimate the $m \times m$ transition matrix $P$ which characterises
this process by setting each entry $p_{ij}$ to be the number of times
state $j$ follows state $i$ in the sequence. The transition activity
of the whole ensemble can be calculated by averaging the TM for each
performer. The matrix can be normalised so that the sum of all
elements is equal to one.

To generate state sequences for our corpus we have used an automated
method to classify logs of touch interaction, at a regular interval,
into nine continuous touch-screen gestures, as described by Martin et
al~\cite{Martin:2015jk}. A Random Forest~\cite{Breiman:2001kx}
classifier\footnote{The Random Forest classifier used was from
  Python's Scikit-learn package\cite{scikit-learn}.} was applied to
5-second windows of touch-data in our corpus to provide chains of
gesture-states for each participant in each session. These chains can
be used to create $9 \times 9$ transition matrices for individual
performers and the whole ensemble in each session.

@@

While the interaction sessions recorded in our corpus were performed
by many different participants with varying experience levels, in
different contexts and on different instrumental setups, the common
interface and method of recording the data allows all the sessions to
be transformed into gesture sequences, and transition matrices. While
these matrices can be used to visualise the structure of individual
performances~\cite{Swift:2014tya}, we wish to extract quantitative
measures from these TMs to compare the large number of sessions in our
corpus.

To do this we calculate two high-level quantities from each session:
``flux'' and ``entropy''. The flux of a sequence is defined as the
ratio of state transitions (e.g. $A \rightarrow B$ where $A \neq B$),
to self transitions (e.g. $A \rightarrow A $). The flux of a TM $P$ is
given by: $\mathrm{flux}(P) = 1 - \mathrm{tr}(P)$ where the trace
$\mathrm{tr}(P)$ is the sum of the diagonal entries of $P$.
Intuitively, flux is a measure of how frequently the participant/group
changes state. Flux returns a value in the interval $[0,1]$ and will
return 0 when participants never change state, and 1 when participants
never stay in the same state for two measurements in a row.

Another useful measure that can be used on a transition matrix is its
entropy, defined in the information-theoretic\cite{Shannon:1948rt}
sense: $H(P) = -\sum_{i,j}p_{ij}\log_2(p_{ij})$. This measure
is small when the matrix is sparse, and largest when each matrix
element is equal. It offers a different perspective on collaborative
interactions than the flux measure by capturing the breadth of the
gestural space explored by the participants throughout the course of
an interaction. Consider the degenerate case of a participant who
alternates between two states for a whole performance: the flux in
this case will be maximal ($\mathrm{flux}(P) = 1$) since there are no
self-transitions, (only $A \rightarrow B$ or $ B \rightarrow A$) even
though the participant has only used a small subset of the state
space. The entropy of this interaction, on the other hand, will be
low. Entropy, therefore, is a measure of how broad the participant's
exploration of the state space is in a given interaction.

@@

\begin{figure}
  \centering
  \includegraphics[width=\linewidth]{figures/sessions-count}
  \caption{Distribution of the sessions in our interaction corpus by
    performance context. Improvisations or compositions were repeated
    several times in rehearsals and study sessions but only performed
    once in concerts.
    \label{fig:count-data}}
\end{figure}

@@

\begin{figure} \centering
  \includegraphics[width=\linewidth]{figures/context-flux-entropy-boxplot}
  \caption{Distributions of flux and entropy values by performance
context. Study and rehearsal sessions had the highest flux for
improvisations. Compositions had lower
flux in rehearsals than performances.
\label{fig:flux-entropy-boxplot}}
\end{figure}

The distribution of flux and entropy with respect to session contexts,
types and instrumentation are shown in Figure
\ref{fig:flux-entropy-boxplot}. A three-way ANOVA procedure was
performed on the data set to find whether the metadata could
significantly predict the entropy and flux values.

All main effects on flux were found to be significant: performance
context ($F(2,86) = 6.28, p < 0.01$), performance type
($F(1,86) = 7.21, p < 0.01$), and instrumentation
($F(1,86) = 20.06, p < 0.001$). Significant interaction effects were
also found for performance context and type
($F(1,86) = 6.18, p < 0.05$), and performance context and
instrumentation ($F(1,86) = 10.48, p < 0.01$).

For entropy, significant main effects were found due to performance
context ($F(2,86) = 12.10, p < 0.001$), and instrumentation
($F(1,86) = 25.86$). The interaction of performance type and
instrumentation was also found to be significant
($F(1,86) = 9.93, p<0.01$).

These tests show that the different interaction styles that might be
present in performance contexts, types and instrumentations are
reflected in the values of flux and entropy of their transition
matrices.

\begin{figure*}[t!]
    \centering
    \begin{subfigure}[t]{0.5\linewidth}
        \centering
          \includegraphics[width=\linewidth]{figures/type-section-flux-entropy}
          \caption{Performances divided by performance type. \label{fig:type-section-flux-entropy}}
    \end{subfigure}%
    ~ 
    \begin{subfigure}[t]{0.5\linewidth}
        \centering
          \includegraphics[width=\linewidth]{figures/context-section-flux-entropy}
          \caption{Sessions divided by performance context. \label{fig:context-section-flux-entropy}}
    \end{subfigure}
    \caption{The flux and entropy of different interaction sessions
      divided into beginning, middle, and end.}
\end{figure*}

\subsection{Breaking performances into sections}

It is clear from these results that flux and entropy can be used to
differentiate between performance styles and types, but can they be
used to understand some of the internal structure of performances? A
classical model for understanding temporal art-forms is the
``beginning, middle, end'' structure~\cite{Aristotle:350rt}. To
investigate how this structure fits our corpus, we can divide the
gestural-sequences of each session into thirds and calculate the
transition matrices given by each section.

Figure~\ref{fig:type-section-flux-entropy} shows that the flux and
entropy of compositions are much more stable with respect to section
than for improvisations. The stability of compositions is to be
expected since performers use the same gestures in each rehearsal and
performance of these written scores, however it may be that different
kinds of compositions could result in different structural
differences. Both measures have a downward trend through sections in
improvisations with Kruskal-Wallis tests showing a significant effect
of section on flux ($\chi^2(2)=5.8, p=0.05$). This could be explained
by performers entering an exploration stage at the start of an
improvisation, where they change gesture frequently to experiment with
different ideas. At the end of an improvisation, performers may change
gesture less frequently while ``winding up'' the performance.

The gestural change in improvisation can be further explored by
dividing the sessions by performance context. Figure
\ref{fig:context-section-flux-entropy} shows this division for
improvisations only. In rehearsals and studies the distribution of
flux and entropy over sections has a similar shape, both measures drop
in the ending section of studies and after the beginning section in
rehearsals. In performances however, flux drops after the beginning
while entropy drops substantially in the ending section. This
indicates a more constrained performance style in this closing
section, where performers transition between fewer gestures in the
final section of performances.

@@

While the interaction session in our corpus have many different
contexts, they are all part of a program of artistic and HCI research
that has evolved over time. Figure \ref{fig:flux-entropy-through-time}
shows the values of flux and entropy for each session in the corpus.
Over the course of this research program, different ensembles and
activities have been emphasised and some of these differences can be
seen as clusters of performances with similar flux and entropy in the
two plots. For instance, the earliest series of rehearsals and
performances had flux between 0.05 and 0.15, while the period of
activity after July 2014 had flux between 0.1 and 0.25. This change
corresponded with an intensive series of study sessions and rehearsals
where the participants developed a more ``fluctuating'' style of
performance. The entropy between these two periods, however, remains
at a similar level, showing the the space of available gestures that
the performers explored was relatively stable.

\begin{figure}
  \centering
  \includegraphics[width=\linewidth]{figures/flux-entropy-through-time}
  \caption{Distribution of flux and entropy values of sessions through time.
    Different performance styles are visible as clusters of similar
    flux regardless of performance context.
    \label{fig:flux-entropy-through-time}}
\end{figure}

@@

In this note we have extended previous work using transition matrices
to understand HCI processes with the application of two simple matrix
measures, flux and entropy, both of which have been demonstrated to
have significant discriminatory power with regards to a corpus of 95
collaborative touch-screen interactions. Not only do these measures
differentiate between different types of sessions and the internal
structure of these sessions, they provide qualitative insight into the
performances as well. For improvised performances, live performances
had the lowest measure in both flux and entropy, while lab-studies had
the highest. This may be due to the increased pressure on performers
when improvising in front of an audience. The internal investigation
of performances session revealed more structural variety in
improvisations than compositions. In response, new works could be
composed that specifically explore different gestural structures in
touch-screen performance.

Data-mining logs of interactions with creative interfaces suggest
further approaches for future work. The macro-gestures that these
processes reveal could be used as the basis for new works of new media
art~\cite{Manovich:2002ly}. Additionally, a wider dissemination of
such log data could better support the replication of findings in HCI
research~\cite{Wilson:2011ve}. Finally, flux and entropy proved useful
in the analysis of our data corpus, but other transition matrix
metrics may be more suitable for different interaction data.

HCI researchers are increasingly concerned with interactions that
happen in-the-wild, where the creative collaborations of many users
may be recorded, but where conducting surveys or interviews is
innapropriate. We suggest that analysing interaction-state transitions
using matrices and simple measures can lead to new insights into HCI
interactions outside the lab.

%\pagebreak
\bibliographystyle{SIGCHI-Reference-Format}
\bibliography{references}

>>>>

%\pagenumbering{arabic}

@@

\title{Mining Interaction Log Data in a Creative Arts Practice using Transition Matrices}

@@

\numberofauthors{3}
\author{%
  \alignauthor{1st Author Name\\
    \affaddr{Affiliation}\\
    \affaddr{City, Country}\\
    \email{e-mail address}}\\
  \alignauthor{2nd Author Name\\
    \affaddr{Affiliation}\\
    \affaddr{City, Country}\\
    \email{e-mail address}}\\
  \alignauthor{3rd Author Name\\
    \affaddr{Affiliation}\\
    \affaddr{City, Country}\\
    \email{e-mail address}}\\
     \alignauthor{4th Author Name\\
    \affaddr{Affiliation}\\
    \affaddr{City, Country}\\
    \email{e-mail address}}\\
}

@@

  In this note we present an analysis of an interaction data archive
  associated with a long-term practice-led research program
  investigating ensemble musical performance practice on touch
  screens. We use transition matrix measures to distinguish between
  sessions of different contexts, and the internal structure of these
  sessions. The data analysis proceeds by first classifying these data
  into higher level gesture types and summarising the performers'
  transitions between these gestures. We find that two
  transition-matrix measures, flux, and entropy, can be used to
  distinguish different performance types and contexts, as well as the
  internal structure of the performances in our corpus.
\end{abstract}

@@

\category{H.5.5.}{Information Interfaces and Presentation (e.g.
  HCI)}{Sound and Music Computing}[Methodologies and techniques]

@@

\begin{figure*}
  \centering
  \includegraphics[width=\linewidth]{figures/three-performance-contexts}
  \caption{Collaborative musical interaction with our touch-screen
    interfaces in two different {\em contexts} and two different {\em instrumentations}: (left to right) rehearsal
    in a studio with iPads only,
    concert performance with iPads and percussion instruments, and concert performance with 
    iPads only.\label{fig:three-performance-contexts}}
\end{figure*}

Modern computing devices and interfaces continue to present
opportunities to support new kinds of creative expression and
collaborative interaction~\cite{Resnick:2005yu}. However, it has been
acknowledged that understanding and evaluating interfaces in creative
contexts can be very challenging~\cite{Shneiderman:2007qv}. This problem is
even more pronounced in artistic \emph{performance}, where subjective
interface evaluation through questionnaires or interviews may be both
practically difficult and culturally inappropriate. However, when an artistic interface
is digital, it becomes possible to log artistic performances and these
logs can be used for curation and
preservation~\cite{England:2014ys}. A question we address is: can this log data
be used to understand and support a
broader creative-practice-led research program?

In this note we will show how an archive of collaborative performance
logs can be subjected to a quantitative analysis based on
transition probabilities between gestural states. We will demonstrate
that two measures on transition matrices, flux and entropy, show
significant differences between different kinds of performances, and
give insight into the internal structure of these performances.

Our archive of interaction data-logs has been accumulated to support a
long-term, practice-led, research program into simple iPad interfaces
for group music-making similar to those described in Martin et
al~\cite{Martin:2014cr}. Over a period of more than two years (April
2013--June 2015), we collected data from group music-making sessions
where the participants used our custom-built, touch-screen apps on
Apple iPad devices. Although several different apps have been included
in our corpus, they shared a common mode of free-form touch
interaction where tapping produced short sounds and swirling or
swiping produced long sounds. Apart from these similarities, the apps
featured a variety of sound palettes, different kinds of visual
feedback, and different kinds of networked interactions (between the
iPads in an ensemble and with a central server).

In all of our 95 interaction sessions, every touch interaction---every
touch down, dragging movement and release---was time-stamped and
logged to disk for later analysis. These sessions included {\em rehearsals} for upcoming concerts, without an
audience, as well as curated {\em concerts}, in front of a live audience, where 
performative goals were paramount. They also included a number of {\em recording sessions}
for the purposes of formally-structured HCI experiments. In the recording
sessions, the performers were also interviewed and surveyed-- something which would have 
been disruptive and inappropriate for the concerts. Survey and interview data were not part of the
corpus. 
Some of the sessions were
iPad-only and others used iPads together with other acoustic
percussion instruments. Although the majority of the sessions were
free-form improvisations, a number of rehearsals and performances
involved composed pieces where the musicians read and performed
percussion notation using the apps.

Some of these session contexts are are shown
Figure~\ref{fig:three-performance-contexts} and other descriptive
statistics about the corpus are shown in Table \ref{corpus-table}.
In our corpus, the median length of sessions is 7m26s (IS THIS STILL THE MEDIAN IGNORING THE
DEMONSTRATION AND DATA GATHERING SESSIONS?) and the median
number of participants per ensemble is four. Each session record is
labelled with metadata about its \emph{context} (concert,
rehearsal, recording), its \emph{type} (improvisation or composition),
and its \emph{instrumentation} (iPads-only, or iPads with acoustic
percussion). The distribution of recorded sessions by context and type
is shown in Figure \ref{fig:count-data}.

@@

\begin{table}
\centering
\begin{tabular}{l|llll}
\hline
            & Total & Min  & Median   & Max     \\ 
\hline
N           & 95    &      &          &         \\
%Record size & 306MB     &         &        &   \\
Length & 13H9M19S & 1M19S & 7M26S & 22M20S \\
Participants& 353   & 2    & 4        & 9        \\
Flux &   & 0.02239 & 0.1445 & 0.3246\\
Entropy &   & 0.7683 & 3.342 & 4.442\\          
\hline
\end{tabular}
\caption{
  Representative statistics from the corpus of musical touch-screen
  interaction sessions described in this note.  Each record in the corpus
  consists of low level touch-data (such as touches begin, touches move, touches end) recorded in CSV format. 
  Many of the participants were
  present in multiple sessions, but this data was not tracked in
  our (deidentified) corpus.\label{corpus-table}}
\end{table}

@@

A useful way of modelling user interfaces is to consider their
theoretical configurations as state vectors and to deploy the methods
of matrix algebra to describe transitions between these states during
an interaction sequence~\cite{Thimbleby:2001kq, Thimbleby:2004fj}. An
inverted approach is to record users' interactions and manually code
interaction states to produce a transition model. This method has been
applied to usability analysis of many applications outside of the creative arts, for example~\cite{Kannampallil:2007fp}
studied resource-planning software for
security applications and used the transition
probability matrix of these state-sequences to draw
conclusions about the nature of interactions with that software. As an example from the creative
arts, protocols obtained from a computer interface for ``live coding''
computer music have been  coded as states and analysed using
transition matrices~\cite{Swift:2014tya} in both the textual (editing) and
musical dimensions, in order to arrive at conclusions regarding
artistic style.

In our corpus of iPad musical performances, logs of low-level touch-data record each performer's micro-gestures. 
Because it has been
established that musical performances on touch-screens can be chuncked into
vocabularies of continuous macro gestures that unfold over
several seconds of touch interaction~\cite{Martin:2014cr}, these logs
 can be thought of as a sequence of such
states~\cite{Pressing:1988uo} for each performer.
If a performers' touch interactions can be classified as belonging
to one of $m$ distinct states, then, over the course of a session, each
participant $X$ generates a length $N$ sequence of these interaction
states $X_n$ ($n = 1, \ldots, N$) which can be thought of as Markov
process.
If there is an (automated) classification process for identifying each
participant's state sequence from their interaction data, then we can
estimate the $m \times m$ transition matrix (TM), $P$, which characterises
this process by setting each entry, $p_{ij}$, to be the number of times
state $j$ follows state $i$ in the sequence. The transition activity
of the whole ensemble can be calculated by averaging the TM for each
performer. The matrix can be normalised so that the sum of all
elements is equal to one.

To generate state sequences for our corpus we have used an automated
method to sample the data at regular intervals and to classify the logs of touch events
into the nine continuous touch-screen gestures identified by Martin et
al~\cite{Martin:2015jk}. A Random Forest~\cite{Breiman:2001kx}
classifier\footnote{The Random Forest classifier used was from
  Python's Scikit-learn package\cite{scikit-learn}.} was applied to
5-second windows of touch-data in our corpus to provide chains of
gesture-states for each participant in each session. These chains can
be used to create $9 \times 9$ transition matrices for individual
performers and the whole ensemble in each session.

@@

While the interaction sessions recorded in our corpus were performed
by many different participants with varying experience levels, in
different contexts and on different instrumental setups, the common
interface and method of recording the data allows all the sessions to
be transformed into gesture sequences, and transition matrices. While
these matrices can be used to visualise the structure of individual
performances, as has been done in ~\cite{Swift:2014tya}, we also wish to extract quantitative
measures from these TMs to compare the large number of sessions in our
corpus.
To do this we calculate two high-level quantities from each session:
``flux'' and ``entropy''. The flux of a sequence is defined as the
ratio of state transitions (e.g. $A \rightarrow B$ where $A \neq B$),
to self transitions (e.g. $A \rightarrow A $). The flux of a TM $P$ is
given by: $\mathrm{flux}(P) = 1 - \mathrm{tr}(P)$ where the trace
$\mathrm{tr}(P)$ is the sum of the diagonal entries of $P$.
Intuitively, flux is a measure of how frequently the participant/group
changes state. Flux returns a value in the interval $[0,1]$ and will
return 0 when participants never change state, and 1 when participants
never stay in the same state for two measurements in a row.

Another, evocative measure that can be defined on a TM is its
entropy, defined in the information-theoretic\cite{Shannon:1948rt}
sense: $H(P) = -\sum_{i,j}p_{ij}\log_2(p_{ij})$. This measure
is small when the matrix is sparse, and largest when each matrix
element is equal. It offers a different perspective on collaborative
interactions than the flux measure by capturing the breadth of the
gestural space explored by the participants throughout the course of
an interaction. Consider a degenerate case of a participant who
alternates between just two states for a whole performance: the flux in
this case will be maximal ($\mathrm{flux}(P) = 1$) since there are no
self-transitions, (only $A \rightarrow B$ or $ B \rightarrow A$) even
though the participant has only used a small subset of the state
space. The entropy of this interaction, on the other hand, will be
low. Entropy, therefore, is a measure of how broad the participant's
exploration of the state space is in a given interaction.

@@

\begin{figure}
  \centering
  \includegraphics[width=\linewidth]{figures/sessions-count}
  \caption{Distribution of the sessions in our interaction corpus by
    session {\em context}. Improvisations or compositions were repeated
    several times in rehearsals and recording sessions but only performed
    once in concerts.
    \label{fig:count-data}}
\end{figure}

@@

\begin{figure} \centering
  \includegraphics[width=\linewidth]{figures/context-flux-entropy-boxplot}
  \caption{Distributions of flux and entropy values by performance
context. Study and rehearsal sessions had the highest flux for
improvisations. Compositions had lower
flux in rehearsals than concerts.
\label{fig:flux-entropy-boxplot}}
\end{figure}

The distribution of flux and entropy with respect to session {\em contexts} (concert,
rehearsal, recording),
{\em types} (improvisation or composition) and {\em instrumentation} (iPads-only, or iPads with acoustic
percussion) are shown in Figure
\ref{fig:flux-entropy-boxplot}. A three-way ANOVA procedure was
performed on the data set to find whether the metadata could
significantly predict the entropy and flux values.

All main effects on flux were found to be significant: session
context ($F(2,86) = 6.28, p < 0.01$), session type
($F(1,86) = 7.21, p < 0.01$), and instrumentation
($F(1,86) = 20.06, p < 0.001$). Significant interaction effects were
also found for session context and type
($F(1,86) = 6.18, p < 0.05$), and session context and
instrumentation ($F(1,86) = 10.48, p < 0.01$).

For entropy, significant main effects were found due to session
context ($F(2,86) = 12.10, p < 0.001$), and instrumentation
($F(1,86) = 25.86$) (WHAT IS THE P VALUE HERE?). There was no main effect for session type although
 the interaction of session type and
instrumentation was found to be significant
($F(1,86) = 9.93, p<0.01$).

These tests show that the different interaction styles that might be
present in the various session contexts, types and instrumentations in our corpus
appear to be descriminated  by values of the flux and entropy of their transition
matrices.

\begin{figure*}[t!]
    \centering
    \begin{subfigure}[t]{0.5\linewidth}
        \centering
          \includegraphics[width=\linewidth]{figures/type-section-flux-entropy}
          \caption{Sessions divided by session type. \label{fig:type-section-flux-entropy}}
    \end{subfigure}%
    ~ 
    \begin{subfigure}[t]{0.5\linewidth}
        \centering
          \includegraphics[width=\linewidth]{figures/context-section-flux-entropy}
          \caption{Sessions divided by session context. \label{fig:context-section-flux-entropy}}
    \end{subfigure}
    \caption{The flux and entropy of different interaction sessions
      divided into beginning, middle, and end.}
\end{figure*}

\subsection{Breaking Sessions into sections}

It appears from our data-mining results that both flux and entropy can be used to
differentiate between session styles and types, but can they be
used to understand some of the internal structure of sessions? Since Aristotle, a
classical model for understanding temporal art-forms is the three-phase
``beginning, middle, end'' structure~\cite{Aristotle:350rt}. To
investigate how this structure fits our corpus, we have divided the
gestural-sequences of each session into three equal sections and have calculated the
transition matrices given by each section.

Figure~\ref{fig:type-section-flux-entropy} shows that the flux and
entropy of compositions are much more stable across the three sections than  are
improvisations. This stability of compositions is to be
expected since performers use the same gestures in each rehearsal and
concert of written scores, however it may be that different
kinds of compositions could result in different structural
differences. Both measures have a downward trend through sections in
improvisations with Kruskal-Wallis tests showing a significant effect
of section on flux ($\chi^2(2)=5.8, p=0.05$). This could be explained
by performers entering an exploration stage at the start of an
improvisation, where they change gesture frequently to experiment with
different ideas. At the end of an improvisation, performers may change
gesture less frequently while ``winding up'' the concert performance.

The gestural change in improvisation can be further explored by
dividing the sessions by session context. Figure
\ref{fig:context-section-flux-entropy} shows this division for
improvisations only. In rehearsals and studies the distribution of
flux and entropy over sections has a similar shape, both measures drop
in the ending section of studies and after the beginning section in
rehearsals. In concerts however, flux drops after the beginning
while entropy drops substantially in the ending section. This
indicates a more constrained performance style in this closing
section, where performers transition between fewer gestures in the
final section of concert performances.

@@

While the interaction sessions in our corpus have many different
contexts, they are all part of a program of artistic and HCI research
that has evolved over time. Figure \ref{fig:flux-entropy-through-time}
shows the values of flux and entropy for each session in the corpus.
Over the course of this data collection, different ensembles and
activities have been emphasised and some of these differences can be
seen as clusters of performances having similar fluxes and entropies in the
two plots. For instance, the earliest series of rehearsals and
concerts, from early 2013 until July 2014, 
had fluxes that were bounded between approximately 0.05 and 0.17.
Just before and after July 2014, fluxes were bounded between 0.1 and 0.26. Finally,
between early and mid 2015, fluxes were bounded between 0.08 and 0.32. Within these 
blocks of performances, distinct clusters can be identified. Clusters in the entropy plots
sometimes show similar trends to the fluxes and sometimes different. The higher
upper bounds of fluxes close to July 2015 relative to those a year earlier may indicate
the development of a more ``fluctuating'' style of performance. In the same period, the bound
of entropies remained more similar indicating that the total space of available gestures continued
to be explored in a relatively stable way.

\begin{figure}
  \centering
  \includegraphics[width=\linewidth]{figures/flux-entropy-through-time}
  \caption{Distribution of flux and entropy values of sessions through time.
    Different performance styles are visible as clusters of similar
    flux regardless of session context.
    \label{fig:flux-entropy-through-time}}
\end{figure}

@@

In this note we have applied transition matrices
to a data corpus of 95 collaborative musical performances using touch screens.
We have applied two simple matrix
measures, flux and entropy, to the data and have shown that
 both of these measures appear to 
have discriminatory power. Not only do these measures
differentiate between different types of sessions, and the internal
structure of these sessions, but they provide qualitative insight into the
performances. For improvised performances, it turns out that our results showed that concerts
had the lowest values of both flux and entropy, while recording sessions had
the highest. This may be due to the increased pressure on performers to take fewer risks
when improvising in front of an audience at a concert and indicates that perhaps rehearsals should 
emphasise the development of such risk-taking. 
Our results for the three phases of performance (beginning, middle, end)
revealed more variation in
improvisations than compositions. As a creative
 response to this finding, 
new works might be
composed that specifically explore and emphasise different gestural sequences across the three phases.

Our data-mining of interaction-logs from creative interfaces suggests
further approaches for future work.
Our flux and entropy measures have proven useful
in the analysis of our data corpus, but many other transition matrix
metrics might be found to be useful for these and different interaction data.
Our assumption of equal intervals for the beginning, middle and end of performances could be 
optimised with a goal to identifying a common, or even optimal, apportionment of time to each phase.  
A wide dissemination of corpuses such as ours could better support the replication of findings in HCI
research~\cite{Wilson:2011ve}. The macro-gestures that these
processes reveal could be used as the basis for ``transcoding'' new works of new media
art~\cite{Manovich:2002ly} from these data corpuses.

%HCI researchers are increasingly concerned with interactions that
%happen in-the-wild, where the creative collaborations of many users
%may be recorded, but where conducting surveys or interviews is
%innapropriate. We suggest that analysing interaction-state transitions
%using matrices and simple measures can lead to new insights into HCI
%interactions outside the lab.

\pagebreak
\bibliographystyle{SIGCHI-Reference-Format}
\bibliography{references}
